---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# recurSurvTests

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/muschellij2/recurSurvTests/graph/badge.svg)](https://app.codecov.io/gh/muschellij2/recurSurvTests)
[![R-CMD-check](https://github.com/muschellij2/recurSurvTests/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/muschellij2/recurSurvTests/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`recurSurvTests` provides transparent, research-oriented reference
implementations for recurrent **gap-time** survival and recurrent competing-risk
analyses. It deliberately keeps three different scientific targets separate:

- marginal recurrent gap-time survival;
- mean cumulative recurrent-event burden over calendar time (not implemented
  here); and
- recurrent cause-specific cumulative incidence.

## Methods

- `wc_surv()` estimates Wang-Chang marginal recurrent gap-time survival using
  the equivalent Luo-Huang weighted risk set.
- `wc_logrank()` implements Luo-Huang's two-sample `G_rho*` statistic. With
  `rho = 0`, it is a recurrent gap-time log-rank analogue; it is not claimed to
  be an exact implementation of Zhao et al. (2020).
- `ss_rcif()` estimates Sivadasan-Sankaran recurrent cause-specific cumulative
  incidence functions (RCIFs).
- `ss_rcif_equal_causes_test()` is a subject-bootstrap Wald reconstruction of
  the Sivadasan-Sankaran/METRON equal-cause RCIF test. It is a weighted CIF
  contrast, not a log-rank test.

## Installation

You can install the development version of recurSurvTests from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("muschellij2/recurSurvTests")
```

## Data convention

Supply one row per observed gap, ordered within subject by `episode` (or already
in episode order). The final row for each subject is a right-censored gap with
`status = 0`; all prior rows have `status = 1`. `gap` is a duration, not a
calendar time. For RCIF analyses, completed events additionally have `cause`.

See `vignette("recurrent-methods", package = "recurSurvTests")` for the
estimands, formulas, implementation details, and a worked example.

## Example

```{r}
#| label: example
library(recurSurvTests)
gaps <- data.frame(
  id = rep(1:4, each = 3),
  episode = rep(1:3, 4),
  gap = c(2, 3, 4, 1, 4, 3, 2, 2, 5, 1, 3, 4),
  status = rep(c(1, 1, 0), 4),
  arm = rep(c("A", "A", "B", "B"), each = 3),
  cause = c("A", "B", NA, "A", "A", NA, "B", "A", NA, "B", "B", NA)
)

wc_surv(gaps, episode = "episode")
wc_logrank(gaps, group = "arm", episode = "episode", rho = 0)$p.value
ss_rcif(gaps, episode = "episode")$curve
```

## References

- Wang MC, Chang SH (1999). *JASA* 94, 146-153.
- Luo X, Huang CY (2011). *Statistics in Medicine* 30, 301-311.
  <https://doi.org/10.1002/sim.4074>
- Sivadasan SM, Sankaran PG (2023). *Statistica* 83, 3-25.
- Sivadasan SM, Sankaran PG (2022). *METRON*.
  <https://doi.org/10.1007/s40300-022-00228-x>
