---
title: "Recurrent Gap-Time Survival, Rank Tests, and Competing Risks Analysis: `recurSurvTests`"
author:
  - name: John Muschelli
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
    code-fold: true
  pdf:
    pdf-engine: xelatex
    documentclass: article
    classoption: [12pt]
    geometry: margin=1in
    number-sections: false
    keep-tex: true
    cite-method: natbib
    biblio-style: unsrtnat
    natbiboptions: numbers,sort
    include-in-header:
      text: |
        \usepackage{booktabs,longtable,threeparttable,lineno}
        \renewcommand{\baselinestretch}{1.45}
        \linenumbers
execute:
  echo: true
  warning: false
  message: false
  cache: false
  error: false
---

```{r}
#| label: setup
#| include: false
here::i_am("vignettes/jss-manuscript.qmd")
repo_root <- here::here()
source(here::here("vignettes", "manuscript-helpers.R"))
if (!requireNamespace("pkgload", quietly = TRUE)) stop("Install pkgload to render the checkout")
pkgload::load_all(repo_root, quiet = TRUE)
options(width = 100)
```

## Abstract

Nonparametric recurrent-event survival curves are available in the statistical
literature, but openly available implementations of tests for differences
between those curves have been limited. `recurSurvTests` provides R
implementations for estimating and comparing recurrent curves on their intended
scales: (i) Wang-Chang marginal recurrent gap-time survival and Luo-Huang
weighted-risk-set rank tests, (ii) the
Peña-Strawderman-Hollander generalized product-limit estimator, (iii) a
separately documented Zhao et al. extended-rank implementation with a
sensitivity variance path, and (iv) Sivadasan-Sankaran recurrent
cause-specific cumulative-incidence functions and their Wald-type comparison.
The package is designed to enable tests whose risk sets, weights, and null
hypotheses correspond to the curve an investigator has chosen to estimate.
Because recurrent gap-time survival, calendar-time event burden, and recurrent
competing-risk incidence are distinct estimands, the software makes those
choices explicit rather than treating their tests as substitutes. This
manuscript documents the methods, data conventions, numerical checks against
prior R implementations where available, and reproducible simulation designs.  We provide a comparison to published tables and also show an example similar to data analysis from polysomnography sleep data analyzing sleep bouts.

**Keywords:** recurrent events, gap time, Wang-Chang estimator, weighted risk
set, product-limit estimator, competing risks, R

## Introduction

Recurrent events occur when a unit can experience an event repeatedly during
observation: hospital admissions, equipment failures, sleep-stage bouts, and
recurrent infections are examples. A frequent implementation problem is that a
method described as a "recurrent survival curve" may not estimate the same
quantity as another method with a similar name. This distinction is especially
important when the number of observed recurrences is associated with a
participant's propensity for short gaps.

This distinction creates a practical inferential problem, not merely a naming
problem. An investigator may plot a recurrent gap-time survival curve and want
to test whether that curve differs by treatment, diagnosis, or visit. A test
for a calendar-time mean cumulative function, a regression coefficient, or a
first-event survival curve instead answers a different question. Thus, the
estimand used for the curve and the estimand targeted by its comparison must be
aligned before a reported p-value has its intended interpretation.

Substantial software already supports recurrent-event regression. For example,
the Andersen-Gill [@andersen1982], Prentice-Williams-Peterson (PWP)
[@prentice1981], and related Cox-model formulations provide score,
Wald, and likelihood-based inference for regression coefficients; mixed-effects
and frailty extensions are available in `coxme` [@coxme] and `frailtypack` [@frailtypack]. These are
important tools when a conditional covariate or frailty effect is the scientific
target. They do not, however, supply a nonparametric global test for equality
of an estimated recurrent gap-time survival curve or tests contrasting such
survival curves across groups. Existing recurrent-curve R
software has also made important contributions: `survrec` [@survrec] includes
Wang-Chang and Peña-Strawderman-Hollander estimators, and `newTestSurvRec` [@newtestsurvrec]
includes recurrent-event curve estimators and rank-test procedures. Their
available estimators and testing paths do not all target the same curve, so a
test cannot be transferred between them solely on the basis of a shared label
such as "recurrent survival."

`recurSurvTests` fills the resulting open-source implementation gap by enabling
tests that correspond to the recurrent curves under analysis. It is not a
general recurrent-event modeling framework. Its primary aim is to let users
estimate a clearly defined recurrent curve and conduct a comparison whose risk
set, weighting, and null hypothesis match that curve--in particular, the
Wang-Chang marginal gap-time survival curve and its Luo-Huang weighted-risk-
set comparison. The package also keeps alternative estimands explicit, rather
than presenting their tests as interchangeable.

### Relationship to `newTestSurvRec::Dif.Surv.Rec()`

The existence of `newTestSurvRec::Dif.Surv.Rec()` does not make the Zhao and
Luo-Huang implementations redundant, but the functions should not be treated
as interchangeable. A source-level inspection of `Dif.Surv.Rec()` shows the
following distinctions:

| Component | `newTestSurvRec::Dif.Surv.Rec()` | Zhao et al. (2020) / `zhao_rank_test()` |
|---|---|---|
| Curve/risk-set path | The function calls `survfitr(..., type = "pe")`, i.e. its displayed test path is built from the Peña-Strawderman-Hollander/GPLE-style pooled gap-record fit. The package also contains a separate Wang-Chang fitter, but that is not the estimator selected inside this function. | Uses Wang-Chang weighted risk and event masses, with each subject's eligible gaps scaled by $1/m_i^*$. |
| Score numerator | At each pooled event time, group-1 events minus the pooled-risk expected group-1 count. | The same broad observed-minus-expected rank-test idea, but with the Wang-Chang weighted counts and the score in Equation 3. |
| Weight families | Provides a broad family (`LRrec`, `Grec`, `TWrec`, `PPrec`, `PMrec`, `PPrrec`, `HFrec`, `CMrec`, `Mrec`). For example, `Grec` uses the total risk-set size and `TWrec` its square root. | Implements the three paper weights LR, GB, and PP. In the paper's notation these are $1$, $n/Y^*(t)$, and the pooled Wang-Chang survival estimate $S^*(t)$, respectively. Thus the label "GB" is not evidence that the `Grec` calculation is the same test. |
| Variance | Uses an event-time variance term based on pooled and group risk-set/event counts, then sums weighted contributions. It does not use Zhao's subject residual vector. | Uses subject-level residuals from Equation 6 and the robust variance $\widehat V^* = \sum_i \hat\epsilon_i^{*2}$ in Equation 7. The pooled-risk and group-specific Equation 6 conventions are retained separately here. |
| Output interpretation | A family of recurrent-curve tests whose default curve path is PSH/GPLE-style. | A test of equality of Wang-Chang marginal recurrent gap-time survival distributions, with a paper-specific robust variance construction. |

Consequently, `Dif.Surv.Rec(type = "LRrec")` is a useful historical recurrent
curve rank test and an important reference comparison, but it is not an exact
implementation of Zhao's Equation 3--7 construction. In particular, matching
the word "log-rank" does not resolve the curve estimator, event weighting, or
within-subject variance choice. The manuscript's numerical checks therefore
compare the curve estimators separately and do not use `Dif.Surv.Rec()` as a
validation oracle for `zhao_rank_test()`.

Three scientific targets are kept separate throughout:

1. Marginal recurrent **gap-time survival**, the target of Wang-Chang and the
   Luo-Huang rank test.
2. Mean cumulative recurrent-event burden on the **calendar-time** scale,
   which is not implemented here.
3. Recurrent cause-specific cumulative incidence for **gap times**, the target
   of the Sivadasan-Sankaran estimators.

This separation avoids calling all recurrent-event comparisons "log-rank
tests." In this manuscript, CIF means cumulative-incidence function, and
RCIF means recurrent cumulative-incidence function. METRON is the journal
*METRON*, where the Sivadasan-Sankaran recurrent-CIF test was published; that
procedure is a weighted
contrast/Wald-type procedure and is not a log-rank test.

## Data convention and software design

All gap-time survival functions accept a long data frame with one row per
observed gap. If an `episode` column is supplied, the package sorts rows within
subject by that column; without it, the existing within-subject row order is
used. Completed gaps have
$\delta_{ij}=1$; the final row is the terminal right-censored gap with
$\delta_{ij}=0$. The observed gap duration is $Y_{ij}$. An optional `episode`
column makes the ordering explicit. This convention permits a participant with
no completed recurrence to contribute a single terminal censored gap.

```{r}
#| label: data-example
library(recurSurvTests)

gaps <- data.frame(
  id = rep(1:4, each = 3),
  episode = rep(1:3, 4),
  gap = c(2, 3, 4, 1, 4, 3, 2, 2, 5, 1, 3, 4),
  status = rep(c(1, 1, 0), 4),
  arm = rep(c("control", "treatment"), each = 6),
  cause = c("A", "B", NA, "A", "A", NA, "B", "A", NA, "B", "B", NA)
)
```

Internal validation enforces this final-censoring convention, nonnegative gap
durations, and subject-level constancy of group and paired-visit labels. The
core helpers turn long data into subject records and form risk/event components
at distinct completed gap times. Keeping these components visible is intended
to make formula-to-code comparison possible.

## Recurrent gap-time survival estimators

### Wang-Chang marginal survival

Let $m_i^*=1$ for a participant with no complete recurrence and otherwise let
$m_i^*$ equal the number of completed gaps. The Wang-Chang weighted risk and
event increments [@wang1999; @luo2011] are

$$
R_{WC}(t)=\sum_i\frac{1}{m_i^*}\sum_{j=1}^{m_i^*} I(Y_{ij}\geq t),
\qquad
dN_{WC}(t)=\sum_i\frac{1}{m_i^*}\sum_{j=1}^{m_i^*}
I(Y_{ij}=t,\delta_{ij}=1).
$$

`wc_surv()` calculates

$$
\widehat S_{WC}(t)=\prod_{u\leq t}
\left\{1-\frac{dN_{WC}(u)}{R_{WC}(u)}\right\}.
$$

The inverse-$m_i^*$ weighting prevents participants with many observed events
from receiving proportionally more influence on the marginal gap distribution.
The implementation uses the weighted-risk-set representation shown by
@luo2011 to be algebraically equivalent to the estimator of @wang1999.

### PSH and pooled Kaplan-Meier

The Peña-Strawderman-Hollander (PSH) generalized product-limit estimator
[@pena2001] uses
all observed gap records with unit weight:

$$
R_{PSH}(t)=\sum_{i,j}I(Y_{ij}\geq t),\qquad
dN_{PSH}(t)=\sum_{i,j}I(Y_{ij}=t,\delta_{ij}=1),
$$

$$
\widehat S_{PSH}(t)=\prod_{u\leq t}
\left\{1-\frac{dN_{PSH}(u)}{R_{PSH}(u)}\right\}.
$$

Therefore `psh_surv()` is numerically identical to an ordinary pooled
Kaplan-Meier calculation on the same long gap records, including terminal
censored gaps. They do not diverge merely because censoring is increased.
They can appear to differ only if "KM" refers to a different data construction,
for example first gaps only, a calendar-time analysis, different treatment of
terminal censoring, or weights. PSH has a renewal/IID gap-model interpretation;
that interpretation is distinct from the Wang-Chang marginal estimand.

In practice, PSH is most defensible when the scientific target is the
distribution of an exchangeable/IID inter-occurrence time under a renewal model
(with independent censoring), or when a pooled gap-record curve is being shown
explicitly as a descriptive event-weighted quantity. It is not a general
replacement for Wang-Chang: subject frailty, dependence of later gaps on
earlier gaps, or informative event counts causes participants with more events
to contribute more gap records and changes the target. In those settings,
`wc_surv()` is preferable for a marginal gap-time survival target, with
subject-level bootstrap inference. PSH can still be reported as a sensitivity
curve, but it should be labeled as the pooled/renewal estimand rather than as a
generic recurrent Kaplan-Meier estimate. The ordinary `survival::survfit()`
curve on the same long records is numerically the same pooled product-limit
calculation; PSH's value is the recurrent-event interpretation and its explicit
renewal assumptions, not a different product-limit formula.

The legacy PSH implementations return an independence-style no-ties
Greenwood/Nelson-Aalen approximation,
$\widehat S_{PSH}(t)^2\sum_{u\leq t}dN(u)/R(u)^2$. It is not generally
appropriate when within-subject gaps are dependent or event counts are
informative. In such settings, clustered uncertainty methods are needed; a
point estimator alone does not solve the inference problem.

### Subject-bootstrap confidence intervals

`recurrent_surv_ci()` resamples complete participant histories and refits
either curve. This is required because recurrent gaps from one participant are
dependent, and because the WC weights are functions of the complete history.
The `ci_method` argument applies to the recurrent-survival curve CI functions
(`recurrent_surv_ci()` and the corresponding curve-contrast workflow):
`"bootstrap"` gives subject-bootstrap percentile intervals or max-$t$ bands,
whereas `"normal"` gives the legacy independence-style pointwise approximation.
The default percentile intervals are pointwise: at a pre-specified time $t$,
they target $P\{S(t)\in CI(t)\}=1-\alpha$. They are the conventional
uncertainty summary for a plotted curve, but do not support a joint claim about
every point of the curve. For that purpose,
`interval = "simultaneous"` uses a subject-bootstrap max-$t$ critical value to
form a band over an explicitly supplied finite time grid. The associated
coverage simulation estimates pointwise and joint coverage separately; its
results remain pending until the configured run has completed.

The default number of bootstrap draws is $B=999$, consistent with common
bootstrap and permutation practice. It is adequate for routine exploration,
but a two-sided 95% percentile interval estimates each 2.5% tail from only
about 25 draws. For final 95% percentile limits or max-$t$ critical values,
the documentation recommends $B\geq 2000$ when computation permits, yielding
about 50 draws in each percentile tail and reducing Monte-Carlo variation.

A pointwise 95% interval has approximately 95% coverage for $S(t_0)$ at one pre-specified gap
time $t_0$. Drawing such intervals at multiple times does not yield 95% joint
coverage of the full curve. The simultaneous max-$t$ band instead targets
$P\{S(t)\ \text{is covered for every }t\in\mathcal T\}\approx 0.95$ for the
user-supplied finite grid $\mathcal T$; it is correspondingly wider. Neither
output is a confidence interval for a survival quantile (an inverse-curve
parameter such as a median gap) or for a between-group curve difference.
Those are separate targets. `recurrent_surv_quantile_ci()` provides a
whole-subject bootstrap percentile interval for a survival quantile; it does
not invert the independence-style normal curve limits because that inversion
does not account for within-subject clustering. A between-group quantile or
curve-difference interval requires its own two-group cluster-resampling
contrast. `recurrent_surv_contrast_ci()` now supplies the latter for
pointwise differences, ratios, or log-ratios and for a finite-grid simultaneous
band, resampling within independent arms or complete paired-visit clusters.
`recurrent_surv_rmst_ci()` supplies a clustered bootstrap interval for the
restricted mean gap time, which remains defined when the median is not reached.

This interface complements, rather than replaces, prior software. The
`survrec` WC and PSH fitters return pointwise standard errors, and its plotting
method draws normal-approximation pointwise limits. `survrec::survdiffr()` also
performs WC or PSH bootstrap resampling for a *selected survival quantile*,
such as a median, and returns a `boot` object for subsequent interval
calculation. Thus bootstrap inference was not absent from earlier R work.
The standard errors returned by the inspected `newTestSurvRec` WC and PSH
fitters are the independence-style Nelson-Aalen/Greenwood approximation,

$$
\widehat{\mathrm{SE}}\{\widehat S(t)\}=
\widehat S(t)\left\{\sum_{u\leq t}\frac{dN(u)}{R(u)^2}\right\}^{1/2}.
$$

Consequently, those values can be used to form a clipped normal pointwise
curve interval, $\widehat S(t)\mathbin{\pm}z_{1-\alpha/2}\widehat{SE}(t)$;
they are neither bootstrap intervals nor simultaneous bands. The same
independence-style form, evaluated with this package's own event and risk
increments, is available with `ci_method = "normal"` to make the legacy-style
capability explicit. It reproduces the PSH standard-error calculation on the
same long records, and should not be relied on for dependent
recurrent gaps or informative event counts, because the displayed variance does
not account for subject-level clustering. In contrast,
`ci_method = "bootstrap"` provides a whole-curve subject-bootstrap workflow:
percentile intervals for fixed times or a max-$t$ band for a finite grid. The
resampling itself is conceptually straightforward; its value here is a uniform
analyst-facing tool that makes the resampling unit, estimator, time grid, and
coverage claim explicit and consistent across WC and PSH analyses.  For a more complex bootstrap procedure (e.g. nested), users must bootstrap the data explicitly and then fit the models.

### Comparing curves

The following executable example compares Wang-Chang, PSH, and an
independently computed pooled Kaplan-Meier curve from `survival` [@survival]
on the same records. It illustrates the estimands, rather than reproducing
a figure from a source paper.

```{r}
#| label: fig-curves
#| fig-cap: "Wang–Chang and PSH/pooled Kaplan–Meier survival on the example records, and their difference."
#| fig-width: 10
#| fig-height: 4
wc <- wc_surv(gaps, episode = "episode")
psh <- psh_surv(gaps, episode = "episode")

step_at <- function(fit, grid) vapply(grid, function(t) {
  ii <- which(fit$time <= t)
  if (length(ii)) fit$surv[max(ii)] else 1
}, numeric(1))
grid <- sort(unique(c(0, wc$time, psh$time)))
km <- survival::survfit(survival::Surv(gap, status) ~ 1, data = gaps)
km_curve <- data.frame(time = km$time, surv = km$surv)
stopifnot(max(abs(step_at(psh, grid) - step_at(km_curve, grid))) < 1e-12)
par(mfrow = c(1, 2))
plot(grid, step_at(wc, grid), type = "s", ylim = c(0, 1),
     xlab = "Gap time", ylab = "Survival")
lines(grid, step_at(psh, grid), type = "s", col = "firebrick", lty = 2)
legend("topright", c("Wang-Chang", "PSH / pooled KM"),
       col = c("black", "firebrick"), lty = 1:2, bty = "n")
plot(grid, step_at(wc, grid) - step_at(psh, grid), type = "s",
     xlab = "Gap time", ylab = "Wang-Chang minus PSH/KM")
abline(h = 0, lty = 2)
```

## Rank tests and paired visits

`wc_logrank()` implements the two-group Luo-Huang weighted-risk-set [@luo2011]
$G_\rho^*$ statistic. At each event time it contrasts weighted observed events
with their group-specific weighted risk-set expectation. With $\rho=0$, it is
the recurrent gap-time log-rank analogue. Its implementation uses
$w(t)=\widehat S_{WC}(t-)^\rho$, where the pooled Wang-Chang survival estimate
is evaluated just before the event time. Thus $\rho=1$ downweights late events
in proportion to the estimated survival remaining, a Peto-Prentice-type
weight; it is not the same as the log-rank weight $w(t)=1$, which gives every
event time equal weight. The variance is built from subject-level contributions
to the score: each subject's observed-minus-expected weighted event increments
are summed across that subject's gaps, preserving their covariance rather than
treating the gap rows as independent.

`zhao_rank_test()` is separate. It implements the extended LR, Gehan-Breslow,
and Peto-Prentice score construction described by @zhao2020. Its default
`pooled_risk` residual is assessed by the simulations below; current results
are diagnostics of this implementation and do not establish byte-for-byte
reproduction of the published tables. The
group-risk denominator printed in Zhao et al.'s Equation 6 is retained as the
explicit `zhao_eq6` sensitivity/reproduction path. Neither function should be
described as an exact substitute for the other.

There is an unresolved implementation ambiguity in the published description.
Equation 6 appears to define the residual denominator using the risk set of
the subject's own group, whereas the pooled-risk construction uses the pooled
weighted risk set. These choices produce different subject residuals and hence
different robust variance estimates even when the score numerator is unchanged.
The package therefore labels them separately (`variance_method =
"pooled_risk"` and `"zhao_eq6"`) and does not silently treat one as the exact
published convention. We have drafted a clarification request to the authors
asking which denominator and finite-sample scaling were used in the published
simulations; until that is resolved, the comparisons below are implementation
diagnostics rather than a claim of exact reproduction.

For two paired visits per participant, `wc_paired_permutation()` randomly swaps visit labels
within participant pairs and recalculates $G_\rho^*$. Its p-value tests the
within-pair label-exchangeability null. The independent-arm analytical p-value
from `wc_logrank()` is returned for context but is not a paired analysis.
This is the package's permutation-based two-group procedure; the K-group
extension currently has an analytical score test (`wc_logrank_k()`), not a
general K-group label-permutation wrapper.

## Recurrent competing risks

`ss_rcif()` implements recurrent cause-specific cumulative-incidence functions
(RCIFs) following @sivadasan2023. The default shared weighting uses the same
subject/event-count denominator for the pooled and cause-specific processes, so
the cause-specific increments add to the pooled increment. The 2023 paper's
printed Equation 15 instead uses a cause-specific $m_{il}$ denominator and an
indicator for subjects with at least two events of that cause. Taken literally,
that expression can make the cause-specific increments nonadditive. The
`cause_weighting = "literal_eq15"` option is retained for sensitivity and
formula-reproduction purposes; it is not the default because additivity is then
not guaranteed.

`ss_rcif_equal_causes_test()` is a subject-bootstrap Wald reconstruction of the
equal-cause RCIF test proposed by Sivadasan and Sankaran in *METRON* (the
journal) [@sisuma2022]. It compares integrated contrasts
$F_l(t)-F(t)/k$ across causes. It is a weighted cumulative-incidence contrast
with a bootstrap covariance estimate, not a log-rank test, and its
reconstruction should not be presented as byte-for-byte author software.

## Validation against prior implementations

Each row compares the same event-time grid and
survival values with the indicated installed package [@survrec;
@newtestsurvrec]. 

```{r}
#| label: tbl-reference-checks
#| tbl-cap: "Numerical comparisons for Wang–Chang and PSH estimators. Missing differences indicate an unavailable reference package."
reference_checks <- manuscript_reference_checks()
reference_display <- reference_checks
reference_display$Max_absolute_difference <- ifelse(
  is.na(reference_checks$Max_absolute_difference), "Not run",
  formatC(reference_checks$Max_absolute_difference, format = "e", digits = 3))
knitr::kable(reference_display)
```

These are software comparisons for the estimators of @wang1999 and @pena2001.
They are not reproductions of the simulation tables in those papers. In
particular, `newTestSurvRec`'s `LRrec` pathway is associated with PSH-style
estimation and is not a numerical reference for the Luo-Huang rank test.

**Published-benchmark placeholder for Wang-Chang and Luo-Huang.**
The curve in @fig-curves, the independent pooled-KM check, and
@tbl-reference-checks provide reproducible estimator checks. Direct numerical
reproduction of the original Wang-Chang and Luo-Huang simulation tables
[@wang1999; @luo2011] remains a separate validation task requiring a matched
design and transcribed published benchmarks. Zhao's results below assess a
different rank-test construction and cannot fill that role. The existing
K-group polysomnography (PSG) simulation is an application-specific extension check, not a
reproduction of Luo-Huang's two-group study.

## Simulation validation

### Zhao published-table comparison {#sec-zhao-validation}

This section compares completed runs with the "**New method**" columns of
Tables 1--3 in @zhao2020. Table 1 reports the SD of the standardized null
statistic; Table 2 reports one-sided type-I error at $\alpha=0.025$; Table 3
reports power. LR, PP, and GB denote log-rank, Peto-Prentice, and
Gehan-Breslow weights. OLR and Jung-Jeong columns are not used as benchmarks
for this implementation because the manuscript evaluates the three Zhao
extended tests, not the ordinary unadjusted log-rank or the separate
Jung-Jeong clustered statistic. Those columns remain useful contextual
comparators in the source paper but would answer a different validation
question; they are not being claimed as reproduced by `zhao_rank_test()`.

`data-raw/zhao-2020-table-simulations.R` defines 120 cells with four baseline
distributions, three heterogeneity ranges, 100 subjects per arm, study end
180 days, and entry uniform on $(0,180)$. Subject gap multipliers follow
$Z\sim U(0.5,1.5)$, $U(0.1,1.9)$, or $U(0.01,1.99)$, labeled low, medium,
and high heterogeneity. The intended numbers of simulated datasets are
100,000 per null cell and 10,000 per power cell. Treatment gaps in Table 3
are multiplied by $\exp(0.25)$ or $\exp(0.5)$; for Weibull shape 2 these
correspond to changes of 0.5 and 1 in $-\log(\lambda)$.

The baseline in the corrected generator match Table 1: exponential rate
$\exp(-4)$, Weibull shape 2 and scale $\exp(4)$, log-normal parameters
$(\mu,\sigma)=(4,0.5)$, and log-logistic survival
$S(t)=\{1+[\exp(-4)t]^2\}^{-1}$. The study-specific generator uses administrative
censoring. Zhao used `survsim::rec.ev.surv()` [@survsim] and discussed an
additional noninformative censoring mechanism without full parameter details
in Section 3. Thus matching baseline distributions does not establish identical
simulation designs.

The apparent Weibull parameter difference is only a parameterization issue:
Zhao's Table 1 writes $S(t)=\exp(-\lambda t^p)$ with $p=2$ and
$\lambda=\exp(-8)$. Since an R Weibull distribution uses
$S(t)=\exp\{-(t/\text{scale})^p\}$, the corresponding scale is
$\lambda^{-1/p}=\exp(4)$, exactly the value used by the generator. The
exponential implementation uses rate $\exp(-4)$, so its mean is $\exp(4)$;
the log-normal uses `meanlog = 4`, and the corrected log-logistic inverse draw
uses scale $\exp(4)$ and shape parameter $0.5$. A 200,000-draw source-level
sanity check gives means approximately 54.6, 48.4, 61.9, and 85.5 days for
the four distributions, respectively, consistent with the paper's stated
approximate means 55, 48, 62, and 86. The remaining differences from Zhao's
published simulations are therefore design differences (notably the
`survsim` generator and incompletely specified censoring mechanism), not an
unintended `exp(4)`/`exp(-4)` scale inversion.

The reference CSV, `data-raw/zhao-2020-published-tables.csv`, was transcribed
from pages 93--95 of the publisher PDF. **Table 2 has a scale ambiguity:** its
caption says "(*100)" but prints integers such as 27 and 25. We interpret these
as 0.027 and 0.025, consistent with Section 3.2's nominal 0.025 level and
interpretation. This is an explicit, unconfirmed correction; both raw printed
values and interpreted probabilities are retained. Table 2 comparisons are
conditional on this interpretation.

```{r}
#| label: zhao-import
zhao <- manuscript_zhao(repo_root)
zhao_status <- manuscript_zhao_status(zhao)
zhao_available <- sum(zhao$available)
zhao_comparable <- sum(zhao$available & zhao$design_eligible)
zhao_discrepant <- sum(zhao$comparison == "discrepant")
zhao_method <- if (zhao_available) unique(zhao$variance_method) else "not available"
```

At this render, **`r zhao_available` of 120 cells** are available, of which
`r zhao_comparable` satisfy the recorded design-eligibility checks.
The variance convention recorded in these files is **`r zhao_method`**.
Availability is assessed from files, not scheduler status. Missing files are
shown as *Pending*; they do not contribute zero rejection rates or enter
comparison summaries. `pooled_risk` and the printed Equation 6 group-risk
convention (`zhao_eq6`) remain distinct; the importer rejects mixtures of
variance conventions and inconsistent settings. The `zhao_eq6` comparison
remains pending unless a separate, consistently configured result directory
is supplied.

The cell-availability and design-eligibility counts are retained in the
reproducibility artifacts, but are not presented as a scientific table in the
manuscript. They are execution metadata rather than results for readers.


#### Monte Carlo uncertainty and current findings

For a rejection proportion $\hat p$ based on $B$ replicates, the Monte Carlo
standard error is $\{\hat p(1-\hat p)/B\}^{1/2}$. Rejection-rate intervals use
the Wilson score interval: with $z=z_{1-\alpha/2}$, center
$(\hat p+z^2/(2B))/(1+z^2/B)$ and half-width
$z\{\hat p(1-\hat p)/B+z^2/(4B^2)\}^{1/2}/(1+z^2/B)$. For empirical SDs in Table 1 we use the
normal-theory approximation $\operatorname{MCSE}(s)=s/\sqrt{2(B-1)}$;
replicate-level fourth moments are unavailable, so this is approximate.
A cell receives an asterisk when

$$
|\widehat\theta_{\mathrm{local}}-\widehat\theta_{\mathrm{paper}}|
>1.96\sqrt{\operatorname{MCSE}_{\mathrm{local}}^2+
             \operatorname{MCSE}_{\mathrm{paper}}^2}+0.0005.
$$

The last term allows half a unit of rounding in the published tables. The
calculation assumes independent simulation runs and uses the paper's replicate
counts. It is an exploratory cellwise discrepancy screen, not an equivalence
test or a multiplicity-adjusted global validation rule. A cell without an
asterisk is not proven equivalent. Nominal calibration is assessed separately
in the exported CSV by exact binomial tests against 0.025, with Holm adjustment
across the 36 planned Table 2 cells.

```{r}
#| label: zhao-current-findings
#| results: asis
#| echo: false
if (!zhao_comparable) {
  cat("**Results placeholder.** No design-eligible completed cells are available. ",
      "The tables retain all published benchmarks and will populate on rerender.\n\n")
} else {
  cat(sprintf("Among the %d comparable completed cells, **%d are discrepant** under the criterion above.\n\n",
              zhao_comparable, zhao_discrepant))
  power <- subset(zhao, table == "table3" & available & design_eligible)
  if (nrow(power)) {
    cat(sprintf("Across %d available power cells, present minus published power ranges from **%+.2f to %+.2f percentage points**. ",
                nrow(power), 100 * min(power$difference), 100 * max(power$difference)))
    cat("These results require investigation of the generator/censoring design and variance convention before claiming published-table reproduction.\n\n")
  }
}
```

The original RDS files retain summary moments, not individual simulated
statistics or failed-p-value counts. Consequently, the null histogram in
Zhao's Figure 1 cannot be reconstructed from this batch. No histogram is
inferred from a mean and SD. The figure below instead compares the reported
quantities directly, while the following three tables retain the distribution,
heterogeneity, alternative, and LR/PP/GB arrangement of Zhao's tables.

```{r}
#| label: fig-zhao-comparison
#| fig-cap: "Published versus study simulation summaries. The dashed line is equality. Bars show simulation 95% Monte Carlo intervals (normal approximation for SDs, Wilson intervals for rejection rates). Paper uncertainty and rounding also enter the discrepancy screen. Only available, design-eligible cells are plotted."
#| fig-width: 12
#| fig-height: 4.2
manuscript_zhao_plot(zhao)
```

```{r}
#| label: tbl-zhao-null-sd
#| tbl-cap: "Comparison with Zhao Table 1: empirical SD of the standardized null statistic at medium heterogeneity. LR = log-rank, PP = Peto-Prentice, and GB = Gehan-Breslow. Parentheses contain simulation Monte Carlo SEs; an asterisk marks a discrepancy."
knitr::kable(manuscript_zhao_table(zhao, "table1"), booktabs = TRUE,
             align = c("l", "l", rep("r", 6)))
```

```{r}
#| label: tbl-zhao-type1
#| tbl-cap: "Comparison with Zhao Table 2: one-sided type-I error probabilities. LR = log-rank, PP = Peto-Prentice, and GB = Gehan-Breslow. Published integers are divided by 1000 as explained in the text. Parentheses contain simulation Monte Carlo SEs; an asterisk marks a discrepancy."
knitr::kable(manuscript_zhao_table(zhao, "table2"), booktabs = TRUE,
             align = c("l", "l", rep("r", 6)))
```

```{r}
#| label: tbl-zhao-power
#| tbl-cap: "Comparison with Zhao Table 3: one-sided power probabilities. LR = log-rank, PP = Peto-Prentice, and GB = Gehan-Breslow. Parentheses contain simulation Monte Carlo SEs; an asterisk marks a discrepancy. The log-time shift is 0.25 or 0.5 for all four distributions."
knitr::kable(manuscript_zhao_table(zhao, "table3"), booktabs = TRUE,
             align = c("l", "l", "r", rep("r", 6)))
```

**Full-run placeholder.** These same chunks will produce the full-data tables
and plot after the remaining RDS files are copied into the input directory.
Rendering never starts the 100,000-replicate jobs. Use a separate directory
and `ZHAO_RESULTS_DIR` to render a different variance convention rather than
pooling results across methods. The completed-cell findings above remain
provisional until the design differences and observed discrepancies are
resolved.

### Polysomnography-bout simulation study

Polysomnography (PSG) records sleep in scored 30-second epochs. For a selected
sleep stage, consecutive epochs assigned that stage can be run-length encoded
into bouts, and the gaps between bouts form recurrent gap times. The supplement
simulation is designed to resemble that structure, with: (i) subject frailty,
(ii) visit effects, (iii) persistent burst/recovery states, (iv) 360--480
minute recordings with terminal censoring, (v) unequal independent-arm sizes
(30 control, 45 treatment), and (vi) paired visits with shared participant
frailty. Under the null, treatment and control gaps have the same distribution.
The corresponding PSG-bout data should not be assumed to satisfy the IID gap
assumption required for a renewal interpretation of PSH; stage dependence,
subject frailty, and bout-state persistence are precisely reasons to prefer the
Wang-Chang estimand and clustered inference there.
Power alternatives multiply treatment gaps by 1.25 (longer bouts) or 0.80
(shorter bouts). The script evaluates $G_0^*$, $G_1^*$, Zhao pooled-risk
sensitivity analyses, and paired label permutations where appropriate.

```{r}
#| label: psg-results
#| results: asis
psg_file <- Sys.getenv("PSG_RESULTS_FILE", file.path(repo_root, "data-raw/psg-validation-results.rds"))
if (!grepl("^(/|[A-Za-z]:)", psg_file)) psg_file <- file.path(repo_root, psg_file)
if (file.exists(psg_file)) {
  psg_results <- readRDS(psg_file)
  stopifnot(is.data.frame(psg_results),
    all(c("design", "alternative", "method", "n_sim", "B", "rejection_rate", "monte_carlo_se") %in% names(psg_results)))
  cat("Available PSG summaries are shown with their actual replicate and permutation counts; small runs are pilots, not full-study validation.\n\n")
  print(knitr::kable(psg_results, digits = 4))
} else {
  cat("**Results placeholder.** No combined PSG result file is available. The planned 1,000-replicate/999-permutation results will be imported on rerender when `data-raw/psg-validation-results.rds` is supplied. Set `PSG_RESULTS_FILE` to select a different result file.\n\n")
}
```

### K-group application-specific pilot

The following table imports the existing K-group output, if available. Its
replicate count is displayed explicitly. It assesses the package's extension
of the weighted-risk-set score to three unequal arms under the PSG-bout
design in `data-raw/wc-logrank-k-validation-simulation.R`. It is not a
published numerical benchmark from @luo2011, and a small pilot does not
establish type-I-error control.

```{r}
#| label: k-group-results
#| results: asis
k_file <- file.path(repo_root, "data-raw/wc-logrank-k-validation-results.csv")
if (file.exists(k_file)) {
  k_results <- read.csv(k_file)
  stopifnot(all(c("scenario", "rejection_rate", "mc_se", "n_sim") %in% names(k_results)))
  print(knitr::kable(k_results, digits = 4))
} else {
  cat("**Results placeholder.** No K-group output is available.\n\n")
}
```

## Use of AI-assisted development

AI assistance was used during development to help locate and summarize
published methods, draft documentation and simulation scaffolding, translate
published formulas into candidate R code, and propose test cases. AI output was
not treated as statistical validation. Candidate implementations were reviewed against the cited formulas. The
executable checks below assess WC and PSH against independent R packages;
these checks do not validate every rank test or competing-risk procedure. 

## Discussion

The main contribution of `recurSurvTests` is not a claim that one estimator or
test fits every recurrent-event question. Rather, it makes the choice visible:
PSH/pooled KM for a renewal/IID gap interpretation, Wang-Chang for marginal
gap-time survival with event-count weighting, Luo-Huang for the corresponding
weighted-risk-set rank comparison, Zhao-style sensitivity analyses, and RCIF
methods for recurrent competing risks. The reference checks connect the new
implementations to legacy software, while the simulation infrastructure makes
remaining empirical questions reproducible rather than implicit.

Future work includes completing the configured Zhao and PSG simulations,
resolving the observed reproduction discrepancies, further validating the
implemented K-group rank-test extension, and comparing nonparametric tests with marginal and frailty regression
models on planned sleep-bout applications.

### What must be run before a complete manuscript render

The ordinary examples, estimator/reference checks, table formatting, and
manuscript rendering are local and do not launch simulations. The remaining
optional outputs have different costs:

| Output | Driver | Recommended execution | Input needed by render |
|---|---|---|---|
| Zhao Tables 1--3 | `data-raw/zhao-2020-table-simulations.R` via `zhao-2020-table-simulations.sbatch` | HPC. Each of 120 tasks contains 100,000 null or 10,000 power replicates; use the corrected log-logistic tasks plus any other missing task IDs. | One task-specific RDS per task in `data-raw/zhao-2020-table-simulations/`; then render locally. |
| PSG-bout study | `data-raw/psg-validation-simulations.R` via its Slurm array | HPC for the planned 10,000 replicates and 999 paired permutations. Combine task RDS/CSV files into `data-raw/psg-validation-results.rds`. | The combined RDS named above. |
| K-group pilot/validation | `data-raw/wc-logrank-k-validation-simulation.R` via its Slurm array | Local for a small pilot; HPC for 20,000 replicates. Combine task outputs before treating it as a full validation. | `data-raw/wc-logrank-k-validation-results.csv` (the current file is a pilot). |
| Curve-interval coverage | `data-raw/curve-ci-coverage-simulation.R` | HPC recommended: 25,000-subject truth plus 1,000 data sets and 999 bootstraps. Verify task partitioning/aggregation before submitting the current array wrapper. | `data-raw/curve-ci-coverage-results.rds` (not currently available). |
| Zhao report aggregation | `data-raw/zhao-2020-validation-report.R` | Local after Zhao RDS files arrive; it does not simulate. | Existing task RDS files. |
| Final manuscript artifacts | `data-raw/reproduce-manuscript.R` | Local; runs Quarto and writes public tables plus internal metadata. | Any optional combined outputs supplied above. |

At present, the Zhao directory is incomplete. The missing task IDs are
`4, 8, 12, 16, 20, 21, 24, 28, 32, 33, 36, 37, 40, 44, 48, 52, 56, 60,
64, 68, 72, 76, 80, 84, 88, 92, 94, 96, 100, 104, 108, 112, 116, 120`.
IDs divisible by four are the corrected log-logistic cells; the other missing
IDs should be rerun with the same driver as their original cell. The validation
report and manuscript intentionally show these as pending rather than treating
them as zero results.

```{r}
#| label: export-reproducibility
#| include: false
artifact_dir <- Sys.getenv("MANUSCRIPT_OUTPUT_DIR", file.path(repo_root, "docs/jss-manuscript"))
if (!grepl("^(/|[A-Za-z]:)", artifact_dir)) artifact_dir <- file.path(repo_root, artifact_dir)
dir.create(artifact_dir, recursive = TRUE, showWarnings = FALSE)
internal_dir <- file.path(artifact_dir, "_internal")
dir.create(internal_dir, recursive = TRUE, showWarnings = FALSE)
write.csv(zhao, file.path(internal_dir, "zhao-comparison.csv"), row.names = FALSE, na = "")
write.csv(zhao_status, file.path(internal_dir, "zhao-completeness.csv"), row.names = FALSE)
write.csv(reference_checks, file.path(artifact_dir, "reference-checks.csv"), row.names = FALSE, na = "")
for (nm in c("table1", "table2", "table3")) {
  write.csv(manuscript_zhao_table(zhao, nm),
            file.path(artifact_dir, paste0("zhao-", nm, ".csv")), row.names = FALSE)
}
write.csv(attr(zhao, "input_manifest"), file.path(internal_dir, "zhao-input-manifest.csv"), row.names = FALSE)
supplement_files <- c(psg_file, k_file)
supplement_files <- supplement_files[file.exists(supplement_files)]
write.csv(data.frame(file = basename(supplement_files),
  md5 = unname(tools::md5sum(supplement_files))),
  file.path(artifact_dir, "supplement-input-manifest.csv"), row.names = FALSE)
source_files <- c("DESCRIPTION", "vignettes/jss-manuscript.qmd", "vignettes/manuscript-helpers.R",
  "vignettes/references.bib", "data-raw/zhao-2020-validation-report.R",
  "data-raw/zhao-2020-published-tables.csv", "data-raw/zhao-2020-published-tables.md",
  "data-raw/zhao-2020-table-simulations.R", "data-raw/zhao-2020-table-simulations.sbatch",
  "data-raw/psg-validation-simulations.R", "data-raw/wc-logrank-k-validation-simulation.R",
  "data-raw/check-manuscript-inputs.R",
  paste0("R/", list.files(file.path(repo_root, "R"), pattern = "[.]R$", full.names = FALSE)))
write.csv(data.frame(file = source_files,
  md5 = unname(tools::md5sum(file.path(repo_root, source_files)))),
  file.path(artifact_dir, "source-manifest.csv"), row.names = FALSE)
png(file.path(artifact_dir, "zhao-comparison.png"), width = 2400, height = 840, res = 200)
manuscript_zhao_plot(zhao)
dev.off()
pdf(file.path(artifact_dir, "zhao-comparison.pdf"), width = 12, height = 4.2)
manuscript_zhao_plot(zhao)
dev.off()
capture.output(sessionInfo(), file = file.path(artifact_dir, "session-info.txt"))
```

```{r}
#| label: session-info
#| echo: false
sessionInfo()
```

## References {.unnumbered}

::: {#refs}
:::
