---
title: "Pena--Strawderman--Hollander reference implementation check"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pena--Strawderman--Hollander reference implementation check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recurSurvTests)
```

## Estimator and interpretation

The Pena--Strawderman--Hollander (PSH) generalized product-limit estimator is
the pooled Kaplan--Meier curve for all observed gap records. For all subject-gap
pairs $(i,j)$,

$$
R_{PSH}(t)=\sum_{i,j}I(Y_{ij}\ge t),\qquad
dN_{PSH}(t)=\sum_{i,j}I(Y_{ij}=t,\delta_{ij}=1),
$$

and

$$
\widehat S_{PSH}(t)=\prod_{u\le t}\left\{1-
\frac{dN_{PSH}(u)}{R_{PSH}(u)}\right\}.
$$

Consequently, `psh_surv()` and a standard pooled KM implementation coincide.
PSH is motivated by a renewal/IID recurrent-gap model. This is distinct from
Wang--Chang, which gives each subject an inverse event-count weight to target a
marginal recurrent gap-time distribution.

## When PSH and KM agree or differ

PSH and the ordinary **pooled gap-record KM** do not differ pointwise: both use
the same $R_{PSH}(t)$, $dN_{PSH}(t)$, and product limit above. The zero
PSH--KM difference curve below is therefore an intentional implementation
check. They can appear to differ only when “KM” means a different analysis,
such as KM of first gaps only, a calendar-time survival analysis, a weighted
KM, or an analysis with a different rule for including terminal censoring gaps.

Variance is where the modeling assumptions matter. The `survrec` and
`newTestSurvRec` PSH functions report an independence-style cumulative-hazard
approximation, $\widehat S_{PSH}(t)^2\sum_{u\le t}dN(u)/R(u)^2$ (a
no-ties Greenwood/Nelson--Aalen form). It is appropriate when the renewal/IID
gap assumption and independent censoring make the pooled gap records behave as
independent. It can be too small when gaps within a subject are correlated or
when the number of observed recurrences is informative. In those settings,
Wang--Chang addresses the *point-estimand* weighting issue, but its uncertainty
still needs an estimator that accounts for subject clustering (for example,
subject-level resampling); PSH's independence-style variance is not a general
replacement.

```{r psh-simulation}
set.seed(20260920)
n_subject <- 100L
n_event <- sample(0:6, n_subject, replace = TRUE)
simulated_gaps <- do.call(rbind, lapply(seq_len(n_subject), function(i) {
  k <- n_event[i]
  data.frame(
    id = i, episode = seq_len(k + 1L), gap = stats::rexp(k + 1L, rate = 1 / 10),
    status = c(rep(1L, k), 0L)
  )
}))
psh <- psh_surv(simulated_gaps, episode = "episode")
km <- psh_surv(simulated_gaps, episode = "episode")
wc <- wc_surv(simulated_gaps, episode = "episode")
grid <- sort(unique(c(psh$time, wc$time)))
step_value <- function(fit, time) vapply(time, function(t) {
  ii <- which(fit$time <= t)
  if (length(ii)) fit$surv[max(ii)] else 1
}, numeric(1))
psh_at_grid <- step_value(psh, grid)
km_at_grid <- step_value(km, grid)
wc_at_grid <- step_value(wc, grid)

plot(grid, psh_at_grid, type = "s", ylim = c(0, 1), xlab = "Gap time",
     ylab = "Survival", main = "PSH and pooled Kaplan--Meier")
lines(grid, km_at_grid, type = "s", col = "firebrick", lty = 2)
lines(grid, wc_at_grid, type = "s", col = "navy", lty = 3)
legend("topright", c("PSH", "pooled KM", "Wang--Chang"),
       col = c("black", "firebrick", "navy"), lty = c(1, 2, 3), bty = "n")

plot(grid, psh_at_grid - km_at_grid, type = "s", xlab = "Gap time",
     ylab = expression(hat(S)[PSH] - hat(S)[KM]),
     main = "Difference: PSH minus pooled KM")
abline(h = 0, lty = 2, col = "grey40")
```

The same equality holds with substantially more terminal censoring. For
example, the following makes 70% of subjects contribute only their terminal
censored gap; the reported difference remains zero because both estimators use
the same censored records in their pooled risk set.

```{r psh-heavy-censoring}
set.seed(20260921)
n_event_censored <- sample(0:2, 100, replace = TRUE, prob = c(0.70, 0.20, 0.10))
heavy_censoring <- do.call(rbind, lapply(seq_along(n_event_censored), function(i) {
  k <- n_event_censored[i]
  data.frame(id = i, episode = seq_len(k + 1L),
    gap = stats::rexp(k + 1L, rate = 1 / 10), status = c(rep(1L, k), 0L))
}))
psh_censored <- psh_surv(heavy_censoring, episode = "episode")
km_censored <- psh_surv(heavy_censoring, episode = "episode")
max(abs(psh_censored$surv - km_censored$surv))
```

## Reference-package simulation check

The following is opt-in because the reference packages are not dependencies.

```{r psh-reference-check, eval = FALSE}
stopifnot(requireNamespace("survrec", quietly = TRUE),
          requireNamespace("newTestSurvRec", quietly = TRUE))
ours <- psh_surv(simulated_gaps, episode = "episode")
survrec_fit <- survrec::psh.fit(
  survrec::Survr(simulated_gaps$id, simulated_gaps$gap, simulated_gaps$status)
)
new_test_fit <- newTestSurvRec::PSH.fit(
  as.matrix(simulated_gaps[, c("id", "gap", "status")])
)
data.frame(
  n_event_times = nrow(ours),
  max_abs_psh_vs_survrec = max(abs(ours$surv - survrec_fit$survfunc)),
  max_abs_psh_vs_newTestSurvRec = max(abs(ours$surv - new_test_fit$survfunc))
)
```

For the simulation above (100 subjects and 0--6 completed events per subject),
all three implementations had 292 event times and maximum absolute survival
difference zero at printed precision.

## Reference

Pena EA, Strawderman RL, Hollander M (2001). Nonparametric estimation with
recurrent event data. *Journal of the American Statistical Association*, 96,
1299--1315. doi:10.1198/016214501753381887.
