---
title: "Subject-bootstrap confidence intervals for recurrent survival curves"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subject-bootstrap confidence intervals for recurrent survival curves}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(recurSurvTests)
```

`recurrent_surv_ci()` supplies subject-level bootstrap uncertainty intervals
for the Wang--Chang (WC) marginal gap-time curve and the
Pena--Strawderman--Hollander (PSH) curve. The resampling unit is the complete
participant history. Resampling individual rows would incorrectly regard a
participant's recurrent gaps as independent and, for WC, would break the
subject-specific event-count weights.

## Relationship to earlier R implementations

Earlier R software provides valuable, but different, uncertainty tools.
`survrec::wc.fit()` and `survrec::psh.fit()` return pointwise standard errors;
the associated plotting method uses them to draw normal-approximation
pointwise limits. `survrec::survdiffr()` additionally creates WC or PSH
bootstrap replicates for a selected survival quantile and returns a `boot`
object, from which users can request a quantile confidence interval. Therefore
the present function does not introduce subject bootstrap resampling as a new
statistical idea. Instead, it makes full-curve subject bootstrap intervals
available with the same call for WC and PSH, states whether the result is
pointwise or simultaneous, and returns the fitted curve and limits in a common
format. The inspected `newTestSurvRec` WC and PSH fitters provide standard
errors but no exported bootstrap interface.

Those legacy standard errors use the independence-style
Nelson--Aalen/Greenwood calculation

$$
\widehat{\mathrm{SE}}\{\widehat S(t)\} =
\widehat S(t)\left\{\sum_{u\le t}dN(u)/R(u)^2\right\}^{1/2}.
$$

They can produce a normal pointwise curve CI, but they do not produce a
simultaneous band and are not generally adequate for correlated recurrent
gaps. `recurrent_surv_ci(..., ci_method = "normal")` exposes this calculation
using this package's own risk increments. It reproduces the PSH calculation on
the same long records; for WC it is an analogous weighted-risk-set
approximation rather than a claim to reproduce another package's variance
code. Its default subject bootstrap is preferable when participant-level
dependence is relevant.

## Pointwise intervals are the usual curve summary

The default is a percentile interval at each event time. A 95% *pointwise*
interval has approximately 95% repeated-sampling coverage at one pre-specified
time. This is the usual uncertainty display alongside a survival curve; it does
not mean that the full curve is covered with probability 0.95.

```{r pointwise}
gaps <- data.frame(
  id = rep(1:12, each = 3),
  episode = rep(1:3, 12),
  gap = rep(c(1, 2, 4), 12) + rep(0:11, each = 3) / 10,
  status = rep(c(1, 1, 0), 12)
)

wc_ci <- recurrent_surv_ci(
  gaps, estimator = "wc", episode = "episode", B = 199, seed = 1
)
head(wc_ci$curve)
```

The returned `curve` has the estimated survival, bootstrap standard error, and
lower and upper limits. `B = 999` or larger is recommended for analysis;
`B = 199` only keeps this vignette quick. The default `B = 999` follows common
bootstrap and permutation practice. For a two-sided 95% interval, however, its
2.5% and 97.5% limits are determined by only about 25 bootstrap draws in each
tail. Use `B = 2000` or more when stable 95% percentile limits or max-t band
critical values are important and runtime permits.

For a legacy-style normal pointwise interval, use the standard-error option.
It is shown here for comparison, not as the default recurrent-event analysis.

```{r normal-ci}
psh_normal_ci <- recurrent_surv_ci(
  gaps, estimator = "psh", episode = "episode", ci_method = "normal"
)
head(psh_normal_ci$curve)
```

## Simultaneous bands answer a different question

If an analysis makes a statement about the whole displayed curve, request a
simultaneous band. The implementation uses the bootstrap distribution of the
largest absolute, standard-error-scaled curve deviation over the requested
times. It therefore controls coverage jointly over that finite time grid, and
is generally wider than pointwise intervals.

```{r simultaneous}
psh_band <- recurrent_surv_ci(
  gaps, estimator = "psh", episode = "episode", interval = "simultaneous",
  B = 199, seed = 1
)
psh_band$critical.value
head(psh_band$curve)
```

For a band restricted to clinically meaningful gap times, pass them explicitly
via `times`. This both specifies the scientific claim and avoids creating a
wide band over poorly supported tail times.

## Which confidence statement is being made?

These intervals are not interchangeable.

| Quantity | Typical output | Coverage statement |
|---|---|---|
| Curve ordinate at one pre-specified time, $S(t_0)$ | Default bootstrap percentile interval | About 95% coverage at $t_0$ |
| Several separately reported curve ordinates | Pointwise intervals at each time | About 95% at each time separately; not 95% for all times together |
| Entire curve over supplied grid $mathcal T$ | `interval = "simultaneous"` max-t band | About 95% coverage of all $S(t)$ for $t \in \mathcal T$ jointly |
| Survival quantile, such as a median gap | `recurrent_surv_quantile_ci()` subject-bootstrap interval | Coverage for an inverse-curve quantity, not for $S(t)$ |
| Difference between two group curves | Not returned by `recurrent_surv_ci()` | Requires a two-group, cluster-resampling design and a specified contrast |

The normal-SE option is only a pointwise approximation. It cannot become a
simultaneous band by drawing many normal intervals, and it can understate
uncertainty when recurrent gaps are correlated. A simultaneous band is useful
for a global visual or scientific claim, but its width means pointwise
intervals are usually preferable for reporting a small number of pre-specified
times.

## Quantile statistics, including the median

The median gap time is the inverse-curve statistic
$\inf\{t:S(t)\le 0.5\}$. It has a different coverage target from a CI for the
survival ordinate $S(t)$. `recurrent_surv_quantile_ci()` resamples complete
participant histories, refits the selected curve, and takes percentile limits
of the resulting median (or other quantile) estimates.

```{r quantile-ci}
median_ci <- recurrent_surv_quantile_ci(
  gaps, estimator = "wc", episode = "episode", survival_prob = 0.5,
  B = 199, seed = 1
)
median_ci[c("estimate", "lower", "upper")]
```

The function deliberately does not invert the normal/Greenwood curve limits
for this purpose: those limits do not account for clustered recurrent gaps.

## Restricted means and two-group contrasts

When the curve does not reach 0.5, the restricted mean gap time
$\int_0^\tau S(t)dt$ remains defined. It and its subject-bootstrap interval are
available through `recurrent_surv_rmst_ci()`.

```{r rmst}
recurrent_surv_rmst_ci(gaps, estimator = "wc", episode = "episode",
                       tau = 3, B = 199, seed = 1)
```

For independent treatment arms, `recurrent_surv_contrast_ci()` resamples
subjects within arm; supplying `pair` instead resamples complete paired visit
clusters. It provides pointwise differences, ratios, or log-ratios, and a
simultaneous band when requested. These are curve-difference estimands, not the
rank-test statistic.

```{r contrast}
gaps$arm <- rep(c("control", "treatment"), each = 18)
contrast_ci <- recurrent_surv_contrast_ci(
  gaps, group = "arm", estimator = "wc", episode = "episode",
  contrast = "difference", B = 199, seed = 1
)
plot_recurrent_surv_ci(wc_ci)
```

## Coverage check

`data-raw/curve-ci-coverage-simulation.R` is an opt-in simulation runner. It
generates recurrent histories with administrative terminal censoring and
unequal observed event counts, obtains a large-population reference curve for
each estimator, and evaluates pointwise and joint coverage separately. The
large-population reference is intentional: under terminal-gap recurrent data,
the pooled PSH curve need not target an IID exponential gap survival function.

```{r coverage, eval=FALSE}
Sys.setenv(RUN_CURVE_CI_COVERAGE = "true", CURVE_CI_N_SIM = "1000",
           CURVE_CI_B = "999")
source("data-raw/curve-ci-coverage-simulation.R")

out <- readRDS("data-raw/curve-ci-coverage-results.rds")
out$pointwise
out$simultaneous
```

The standard analysis should report pointwise intervals for fixed-time curve
estimates and use simultaneous bands only when making a joint, whole-curve
claim. The simulation output must be reviewed with Monte-Carlo standard errors
before claiming nominal coverage for a chosen setting.
