---
title: "Recurrent gap-time survival and competing-risk methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recurrent gap-time survival and competing-risk methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(recurSurvTests)
```

## Scope and data structure

Recurrent-event methods are often described with similar language while
targeting different quantities. This package is for **gap-time** analyses: if
subject $i$ has observed gap durations $Y_{i1}, \ldots, Y_{im_i}$, each row of
the input represents one such gap. The final row is the right-censored gap
(`status = 0`); preceding rows are completed events (`status = 1`). The optional
`episode` column specifies their within-subject ordering.

This differs from a mean cumulative function, which accumulates events on a
calendar-time scale. The RCIF functions below also use recurrent gap times, but
target cause-specific incidence rather than a survival curve. Thus a result
from `wc_logrank()` should not be interpreted as a test of a calendar-time mean
cumulative function or an RCIF.

## Relationship to existing R packages

The Wang--Chang and Pena--Strawderman--Hollander curves are established
estimators; `newTestSurvRec` also provides recurrent-curve machinery for them.
The corresponding functions in this package (`wc_surv()` and `psh_surv()`) are
transparent reference components, not claims of new estimators. `reda` and
`mets` provide mean cumulative recurrent-event functions and tests for that
calendar-time target. Those methods are useful complements, but they do not
duplicate the gap-time Luo--Huang/Zhao rank tests or the Sivadasan--Sankaran
recurrent-CIF methods documented below. A source-level comparison is recorded
in `docs/related-software.md` in the repository.

For a calendar-time question about the expected number of events by time `t`,
the established choices are `reda::mcfDiff()` or
`mets::recurrent_marginal()` and `mets::test_logrankRecurrent()`. Those tools
also cover AUC-style comparisons and terminal-event settings. This package is
for the complementary gap-time question: comparison of the marginal
distribution of recurrent gaps, or comparison of recurrent cause-specific
incidence curves.

```{r data}
gaps <- data.frame(
  id = rep(1:4, each = 3),
  episode = rep(1:3, 4),
  gap = c(2, 3, 4, 1, 4, 3, 2, 2, 5, 1, 3, 4),
  status = rep(c(1, 1, 0), 4),
  arm = rep(c("A", "A", "B", "B"), each = 3),
  cause = c("A", "B", NA, "A", "A", NA, "B", "A", NA, "B", "B", NA)
)
```

## Wang-Chang marginal gap-time survival

Let $m_i^*$ be the number of gap records used for subject $i$ in the
weighted-risk-set representation. At gap time $t$, the package calculates

$$
R(t) = \sum_i \frac{1}{m_i^*}\sum_j I(Y_{ij} \geq t), \qquad
dN(t) = \sum_i \frac{1}{m_i^*}\sum_j I(Y_{ij}=t,\ \delta_{ij}=1).
$$

The estimated hazard increment is $d\widehat\Lambda(t)=dN(t)/R(t)$ and the
Wang-Chang-equivalent survival estimate is the product

$$
\widehat S(t)=\prod_{u\leq t}\{1-d\widehat\Lambda(u)\}.
$$

In code, `.subject_records()` validates and turns long rows into one record per
subject, `.event_grid()` finds distinct completed gap times, and
`.wrs_components()` constructs the subject-by-time risk and event matrices.
`wc_surv()` sums those matrices and applies the product-limit update.

```{r wang-chang}
wc_surv(gaps, episode = "episode")
```

## Luo-Huang recurrent rank test

For two subject-level groups, `wc_logrank()` applies the Luo-Huang weighted
risk-set $G_\rho^*$ statistic. At each event time it contrasts the observed
weighted events in group 1 with its weighted risk-set expectation, using
$w(t)=\widehat S(t-)^\rho$. `rho = 0` is the recurrent gap-time log-rank
analogue and `rho = 1` is the Peto-Prentice analogue. The variance is formed
from subject-level martingale-style contributions, retaining within-subject
dependence.

```{r rank-test}
rank_fit <- wc_logrank(gaps, group = "arm", episode = "episode", rho = 0)
rank_fit[c("method", "interpretation", "G", "se", "p.value")]
```

This is Luo-Huang's implementation, not the separate Zhao et al. (2020)
score construction. The `zhao_rank_test()` section below keeps those methods
distinct and reports the unresolved Equation 6 variance convention explicitly.

## Zhao extended rank tests and simulation checks

`zhao_rank_test()` is separate from `wc_logrank()` and implements the extended
LR, GB, and PP score construction in Zhao et al. (2020). Its
default `variance_method = "pooled_risk"` has subject residuals that sum
exactly to the score and is covered by the package's null simulation tests.
`"zhao_eq6"` follows the group-specific denominator printed in Equation 6 as a
sensitivity/reproduction implementation. This distinction is explicit because
the paper's published numerical tables cannot be regenerated exactly without
its original simulation software.

```{r zhao-rank}
zhao_rank_test(gaps, group = "arm", episode = "episode",
               variance_method = "pooled_risk")
```

The PSG scenario treats a recording as 8 hours (480 minutes), with no staggered
entry, and scales a chosen gap distribution to target roughly 200 completed
events per participant. Its default `Z ~ U(0.75, 1.25)` avoids the extreme
event-count dispersion in the paper's `U(0.1, 1.9)` stress scenario.

```{r zhao-psg, eval = FALSE}
zhao_rank_simulation(n_sim = 1000, n_per_group = 100, scenario = "psg_200",
                     variance_method = "pooled_risk", seed = 1)
```

## Recurrent competing risks

`ss_rcif()` computes a pooled recurrent hazard and cause-specific increments.
With subject weight $a_i$, the default shared weighting uses $a_i/m_i^*$ for
both pooled and cause-specific event processes. For cause $l$,

$$
d\widehat F_l(t)=\widehat S(t)\,
\frac{dG_l(t)}{R(t)}, \qquad
\widehat F_l(t)=\sum_{u\leq t}d\widehat F_l(u).
$$

The default `cause_weighting = "shared"` maintains pooled/cause increment
additivity. `"literal_eq15"` reproduces the printed Equation 15 denominator as
a sensitivity option; its increments can be non-additive and should be checked
through `additivity_max_abs_dG`.

```{r rcif}
rcif <- ss_rcif(gaps, episode = "episode")
rcif$curve
rcif$additivity_max_abs_dG
```

The optional `ss_rcif_equal_causes_test()` integrates contrasts of
$F_l(t)-F(t)/k$, uses Helmert contrasts, and estimates covariance by resampling
subjects. It is a bootstrap Wald reconstruction of the METRON equal-cause RCIF
test, not a log-rank test. For real analyses use a substantially larger `B`
(for example, 999 or more) than would be appropriate for a quick example.
